Six steps to a better relationship with your future self.

### 2011-03-25 workflow.Rnw
More work

\documentclass[12pt]{article}
\usepackage{fancyvrb,natbib,url,array,dcolumn,float,booktabs,listings}
\usepackage[colorlinks=TRUE]{hyperref}

@@

\begin{document}
\maketitle

<<initialize,echo=FALSE,results=hide>>=
  ##First, just setup the \R environment:
  options(SweaveHooks=list(
            fig=function(){
              par(mar=c(3.5, 3, 1.1, 0),
                  pty="s",
                  mgp=c(1.5,0.5,0),
                  oma=c(0,0,0,0))},#,
            echo=function(){options(continue=" ") ##Don't show "+" prompts,
                            options(prompt=" ")}
            ),
          digits=3,
          scipen=8,
          width=100,
          error=function(){options(prompt="> ",continue="+ ");NULL}
          )

@@

Right? Wrong? If the collaborators find themselves answering ``wrong''
to the questions posed here then reproducing, updating, or changing
the analyses will take a lot of time. If reproduction is hard to do,
then the reputations of the scholars will suffer and, more
importantly, the UN will be unable to implement a master plan for
world peace. This essay provides some suggestions for practices which
will make such reproduction occur much more easily and quickly in the
event that famous papers require special scrutiny. Specifically, this
piece aims to amplify some of what we already ought to know from
\citet{nagler1995coding} and \citet{king1995replication}, and to add
to some of those ideas given current practices, platforms, and
possibilities.

%%% This data frame will be used throughout the analyses.
%%% Putting it up top, so we can use it in the comments section.
<<read.data,echo=FALSE,results=hide>>=
##  This read.csv line is slightly more complex than needed because of
##  the particular encoding used for characters in the source file.
dem.nations.df<-read.csv(url("http://www.hks.harvard.edu/fs/pnorris/Data/Democracy%20CrossNational%20Data/Democracy%20Crossnational%20Data%20Spring%202009.csv",encoding="ISO_8859-1"),as.is=TRUE)
row.names(dem.nations.df)<-dem.nations.df$Natabrv ##make the row names more helpful
good.df<-na.omit(dem.nations.df[,c("Gini2004","protac2000","Nation","meanpr")])
@

@@

Using R, for example, I might specify that the file \Verb+fig1.pdf+
was produced by the following commands in a file called
\Verb+fig1.R+.\footnote{The command \Verb+please-plot+ and some other
  R functions used in this essay come from the
  \Verb+MayIPleaseDoStatistics+ package which emphasizes politeness in
  data analysis. Functions like \Verb+please-plot+ can
  be blocked and more polite versions such as \Verb+may-I-please-have-a-plot+ can be required using
  \Verb+options(politeness=99)+ }

\begin{Verbatim}[fontsize=\footnotesize]
thedata<-read.csv("Data/thedata-15-03-2011.csv") ## Read the data
pdf('fig1.pdf') ## begin writing to the pdf file
please-plot(outcome by explanatory using thedata. red lines please.)
please-add-a-line(using model1)
## Note to self: the following is no different from model1
## model2<-please-fit(outcome by explanatory+explanatory^2 using thedata
## summary(abs(fitted(model1)-fitted(model2)))
dev.off() ## stop writing to the pdf file
\end{Verbatim}

Now, in the future if I wonder how ``that plot on page 10'' was
created, I will know: (1) ``that plot'' is from a file called
\Verb+fig1.pdf+ and (2) \Verb+fig1.pdf+ was created in
\Verb+fig1.R+. In a future where R still exists, if I wanted to change
the figure, I could edit the commands quickly and easily to do so
(since R commands are written in
\href{http://en.wikipedia.org/wiki/Plain_text}{plain text}, and plain
text is a format that will be around as long as computer programmers
write computer programs). In a future where R does not exist, I will
at least be able to read the plain text R commands and use them to
write code in my new favorite statistical computing
language.\footnote{A most, since R is open source, I will be able to
  download R, download an old fashioned open-source operating systems
  (like Ubuntu 10), and run the old-fashioned statistical computing
  environment in the old-fashioned operating system in a virtual
  machine on my new-fashioned actual machine.}

\paragraph{Step 1} Know the provenance of your results so that your
future self or current collaborators can quickly and easily reproduce
your work (and make changes).% --- to know what
% commands created which figures and tables.

\section{No data analyst is an island for long.}
Data analysis involves a long series of decisions. Each decision
requires justification.  Some decisions will be too small and
technical for inclusion in the paper itself. These need to be
documented in the code itself \citep{nagler1995coding}. Paragraphs and
citations in the article itself will justify the most important
decisions.

\subsection{Code to communicate: Comment your code.}
Comments --- unexecuted text inside of a script --- are a message to
collaborators (including your future self) and other consumers of your
work. In the above code, I used comments to explain the lines to
readers unfamiliar with R and to remember that I had tried a different
specification but decided not to use it because adding the squared
term did not really change the predictions.\footnote{R considers text
  marked with ``\#'' as a comment.}

@@

% <<reg1,echo=TRUE,results=verb>>=
% ##  Repeat the regression run for Table~\ref{tab:protest}
% ##  Do political rights predict protest reporting?
% lm1 <- lm(protac2000 ~ I(Gini2004/100) + meanpr, data=good.df)
% ##  Q: Was it worthwhile controlling for political rights?
% lm2 <- lm(protac2000~I(Gini2004/100),data=good.df)
% coef(lm2)[2]-coef(lm1)[2]
% ##  A: Yes (around 3/4 an act of protest difference)
% @

Messages left for your future self (or near-future others) help
retrace and justify your decisions. If you ever need to revist
them as the work moves from seminar paper to conference paper to
poster back to paper to dissertation and onwards. Notice one another
benefit of coding for an audience: we learn by teaching. By assuming
that others will look at your code, you will be more likely to write
clearer code, or perhaps even to learn more about what you are doing
as you do it.

Comment liberally. Comments are discarded when R runs analysis or
LaTeX turns text into a pdf, so only those who dig into your work will
see them.

\subsection{Programming can be literate.}

Sometimes simple comments in the main command file are not enough:
imagine if the authors had arrived at the bootstrap only after running
simulations to assess the coverage of that procedure compared to a
profiled likelihood based approach. Rather than include time consuming
simulations in the body of their main Sweave document, they might have
written some memos that discuss the tradeoffs and implementation
decisions involved in such a comparison. In the main text they would
have just read in the results of the simulation study with an
appropriate comment:

<<bsreg1,echo=TRUE,results=verb,eval=FALSE>>=
## We assessed the coverage of the bootstrap intervals against 
## direct inspection of the likelihood profile in the file
## Memos/shouldwebootstrap.Rnw which produced bsresults.rda

##Here we read in the results of the bootstrap procedure:
load("Memos/bsresults.rda") ## using a compressed binary format here for speed
@

Now that memo itself (or perhaps the main document) probably involves
sentences and paragraphs explaining an analysis. Literate programming
or literate data analysis happens when the code of the analysis is
mixed side by side with the paragraphs explaining, interpreting,
foreshadowing, discussing the analysis.\footnote{This document is
  written in such a way using \LaTeX and R code mixed together in what
  is known as an Sweave file \citep{Leis:2005}. You can download and
  view and edit this document as a plain text file from
  \url{https://github.com/jwbowers/workflow}.}

If comments are guideposts, literate programs are full fledged
travelogues. Why might one care to use a literate programming
practice?

Say you run some command, discover something new (or confirm something
old) and then make a claim backed up by table or a figure. You produce
a nice little report on your work for use in discussions of the
working group. The report itself is a pdf file or some other format
which emphasizes reading.\footnote{like html, postscript, dvi or even
  the wordprocessor formats like rtf or doc or odt, or the web formats
  like html.}  Eventually we want to use pieces of that report
(tables, graphs, paragraphs) in a publishable paper. It would be a
waste of time if you had to re-create those analyses by pointing,
clicking, copying, or pasting. It would also be a waste of time if you
couldn't remember why you decided to use a quadratic instead of a
cubic term and had to re-create that argument. So, we might use
literate programming because we don't want to waste time.

For example, in \S~\ref{sec:data-analys-comp}, I suggested that we
know where ``that plot on page 10'' comes from by making sure we had a
``fig1.pdf'' file produced from a clearly commented plain text file
called something like ``fig1.R.'' If one were using Sweave, one could
even more easily have written the following to produce the desired
figure.

\begin{Verbatim}[fontsize=\footnotesize]
\begin{figure}[H]
\begin{center}
 <<fig1plot,echo=FALSE,fig=TRUE,width=3,height=3>>=
 par(bty="n",xpd=TRUE,pty="s",tcl=-.25)
 ##  Make a scatterplot of Protest by Inequality
 with(good.df,plot(Gini2004/100,protac2000,
                   xlab='Gini Coefficient 2004 (UNDP)',
                   ylab='Mean Protest Activities\n(World Values Survey 1980-2000)',
                   cex=.8))

 ##  Label a few interesting points
 with(good.df[c("EGY","JOR","USA","SWE","CHL"),],
      text(Gini2004/100,protac2000,labels=Nation,srt=0,cex=.6,pos=3,offset=.1))
 @
\caption{Protest activity by income inequality \citep[from][]{norris2009data}.}\label{fig:giniprot}
\end{center}
\end{figure}
\end{Verbatim}

<<fig1code,echo=FALSE,fig=FALSE>>=
##  Make a scatterplot of Protest by Inequality
with(good.df,plot(Gini2004/100,protac2000,
                  xlab='Gini Coefficient 2004 (UNDP)',
                  ylab='Mean Protest Activities\n(World Values Survey 1980-2000)',
                  cex=.8))

##  Label a few interesting points
with(good.df[c("EGY","JOR","USA","SWE","CHL"),],
     text(Gini2004/100,protac2000,labels=Nation,srt=0,cex=.6,pos=3,offset=.1))
@

\begin{figure}[H]
  \begin{center}
<<fig1plot,echo=FALSE,fig=TRUE,width=3,height=3>>=
par(bty="n",xpd=TRUE,pty="s",tcl=-.25)
<<fig1code>>
@
    \caption{Protest activity by income inequality \citep[from][]{norris2009data}.}\label{fig:giniprot}
  \end{center}
\end{figure}

By using \Verb+\label{fig:giniprot}+, I do not need to keep track of
the figure number. Nor do I need a separate \Verb+fig1.R+ file or
\Verb+fig1.pdf+ file.  Tables and other numerical results are also
possible to generate within ``code chunks'' in the source code of a
scholarly paper. Those who view the code for this essay will see how
Table~\ref{tab:protest} was also generated directly from a regression
object.\footnote{See [cite to Beck TPM] for the inspiration for this
  particular presentation of a linear model.}

<<maketable, echo=FALSE, results=hide>>=
  lm1<-lm(protac2000~I(Gini2004/100)+meanpr,data=good.df)  #  Run the regression

@@

library(xtable)
thextab<-xtable(themat,align=c("l",rep(".",4)),
                caption="People living in countries with unequal
  income distributions report less protest activity to World Values
  Survey interviewers than people living in countries with relatively
  more equal income distributions, adjusting for average political
  rights as measured by Freedom House 1980--2000. Data from \\citep{norris2009data}.",label="tab:protest")
@

<<printtable,echo=FALSE,results=tex>>=
  print(thextab,include.colnames=FALSE,
        hline.after=NULL,
        table.placement="H",
        add.to.row=list(
          pos=list(0,3),
          command=c(
            "\\toprule \n &
             \\multicolumn{1}{c}{Coef} &
             \\multicolumn{1}{c}{Std. Err.} &
             \\multicolumn{2}{c}{95\\% CI} \\\\ \\cmidrule(r){2-5} ",
            paste("& \\multicolumn{4}{c}{n: ",n,",",
                  "   resid.sd:",signif(est.sigma),
                  ", R$^2$:",r2,"} \\\\ \\bottomrule",sep="")
            )
          ))
@

If your workflow does not involve \LaTeX and R, you can still
implement some of the principles here (and if you use OpenOffice or
LyX and R you can use \Verb+odfWeave+ do directly do literate data
analysis in a WYSIWYG environment).\footnote{A quick Google search of
  ``Sweave for Stata'' turned up lots of resources for literate
  programming with Stata.}

Literate data analysis is not the same as Sweave, even if Sweave is a
nice implementation.\footnote{See
  \url{http://cran.r-project.org/web/views/ReproducibleResearch.html}
  for many of the different approaches to literature programming for
  R.} For example, the person most often credited with
these ideas used plain \TeX as the ``document formatting'' language
and Pascal as the programming language [cite]. See [X] for an example
of using the Emacs org-mode as literate programming. And, one could
imagine a Word or HTML file in which code chunks are inserted by
commented out and run one at a time (or perhaps extracted by a clever
script from the XML source file of a modern Microsoft Word Document)
--- or an HTML file in which hyperlinks are inserted to plain text
files containing the relevant code chunks.

\paragraph{Principle} Keep in mind the distinction between the
``source code'' of a document (i.e. what computation was required to
produce it) and the visible, type-set page image. Page images are
great for reading, but not great for reproducing or collaborating. The
source code of any document exchanged by the group must be available
and executable.

@@

\subsection{File names send messages to your future self.}

Name your files with evocative and descriptive names. Your
collaborators are less likely to call you at midnight asking for help
if your files are named ``inequality-and-protest-analyses.R'' than if
your files are called ``temp9'' or
``supercalifragilisticexpialidocious.'' (Note the use of the extension
.R to tell us that the file contains R commands. Use extensions like
this as a standard practice to help you and your computer get along.)

\paragraph{Principle} Comments and other documentation enable you to
learn through teaching, code efficiently, and avoid duplicating your
effort. Literate programming also enables you to avoid the kinds of
errors which occur when you re-type numbers in to tables or
graphic-producing systems or even into a document itself. Finally, and
most importantly, a document that can be ``run'' to reproduce all of
the results is a document that can more effectively spur discussion
and learning and cumulation of research as people no longer need spend
weeks attempting to reproduce research on which they desire to build.

@@

\paragraph{Principle} We should know data where the data came from and
what operations were performed on which set of data.

@@

\paragraph{Principle} Writing is rewriting. Thus, all writing involves
versions. When we collaborate with ourselves and others we want to
avoid clobbering and we want to enable graceful reconciliation of
rewriting. One can do these things with formal systems of software
(like subversion, git, etc...) or with formal systems of file naming,
file comparing, and communication or, even better, with both.

@@

\paragraph{Principle} You cannot forsee all of the ways that your code
could be used, but you can at least make sure it does what it is
supposed to do in your particular case.

@@

We all always collaborate. Many of us collaborate with groups of
people at one moment in time as we race against a deadline. All of us
collaborate with ourselves over time.\footnote{What is a reasonable
  time-span for which to plan for self-collaboration on a single idea?
  Ask your advisers how long it took them from idea to dissertation to
  publication.}  The time-frames over which collaboration are
required --- whether among a group of people working together or
within a single scholar's productive life or probably both --- are
much longer than any given version of any given software will easily
exist. \href{http://en.wikipedia.org/wiki/Plain_text#Usage}{Plain
  text} is the exception. Thus, even as we extol version control
systems, one must have a way to ensure future access to them in a form
that will still be around when sentient cockroaches finally join
political science departments (by then dominated by cetaceans after
humans are mostly uploads).\footnote{The arrival of the six-legged
  social scientists revives Emacs and finally makes Ctrl-a Ctrl-x
  Esc-x Ctrl-c a \href{http://kieran.healy.usesthis.com/}{reasonable
    key combination}.}

>>>>

\documentclass[12pt]{article}
\usepackage{fancyvrb,natbib,url,array,dcolumn,float,booktabs,listings}
\usepackage[colorlinks=TRUE]{hyperref}
\usepackage{microtype}

@@

\begin{document}
\frenchspacing
\VerbatimFootnotes
\maketitle

<<initialize,echo=FALSE,results=hide>>=
  ##First, just setup the \R environment:
  options(SweaveHooks=list(
            fig=function(){
              par(mar=c(3.5, 3, 1.1, 0),
                  pty="s",
                  mgp=c(1.5,0.5,0),
                  oma=c(0,0,0,0))},#,
            echo=function(){options(continue=" ") ##Don't show "+" prompts,
                            options(prompt=" ")}
            ),
          digits=3,
          scipen=8,
          width=50,
          error=function(){options(prompt="> ",continue="+ ");NULL}
          )

@@

Right? Wrong? If the collaborators find themselves say ``Wrong'' in
answer to the questions posed here then reproducing, updating, or
changing the original analyses will take a lot of time. If
reproduction is hard to do, then the reputations of the scholars will
suffer and, more importantly, world peace will have been delayed. This
essay provides some suggestions for practices which will make such
reproduction occur much more easily and quickly in the event that
famous papers require special scrutiny. Specifically, this piece aims
to amplify some of what we already ought to know
\citep{king1995replication,nagler1995coding}, and to add to some of
those ideas given current practices, platforms, and possibilities.

%%% This data frame will be used throughout the analyses.
%%% Putting it up top, so we can use it in the comments section.
<<read.data,echo=FALSE,results=hide>>=
##  This read.csv line is slightly more complex than needed because of
##  the particular encoding used for characters in the source file.
dem.nations.df<-read.csv(url("http://www.hks.harvard.edu/fs/pnorris/Data/Democracy%20CrossNational%20Data/Democracy%20Crossnational%20Data%20Spring%202009.csv",encoding="ISO_8859-1"),as.is=TRUE)
row.names(dem.nations.df)<-dem.nations.df$Natabrv ##make the row names more helpful
good.df<-na.omit(dem.nations.df[,c("Gini2004","protac2000","Nation","meanpr")])
good.df$gini04<-good.df$Gini2004/100
good.df$protac00<-good.df$protac2000
@

@@

Using R \citep{R:2011}, for example, I might specify that the file \Verb+fig1.pdf+
was produced by the following commands in a file called
\Verb+fig1.R+.\footnote{The command \Verb+please-plot+ and some other
  R functions used in this essay come from the
  \Verb+MayIPleaseDoStatistics+ package which emphasizes politeness in
  data analysis. Functions like \Verb+please-plot+ can
  be blocked and more polite versions such as \Verb+may-I-please-have-a-plot+ can be required using
  \Verb+options(politeness=99)+ }

\begin{Verbatim}[fontsize=\footnotesize]
thedata <- read.csv("Data/thedata-15-03-2011.csv") ## Read the data
pdf('fig1.pdf') ## begin writing to the pdf file
please-plot(outcome by explanatory using thedata. red lines please.)
please-add-a-line(using model1)
## Note to self: a quadratic term does not add to the substance
## model2<-please-fit(outcome by explanatory+explanatory^2 using thedata
## summary(abs(fitted(model1)-fitted(model2)))
dev.off() ## stop writing to the pdf file
\end{Verbatim}

Now, in the future if I wonder how ``that plot on page 10'' was
created, I will know: (1) ``that plot'' is from a file called
\Verb+fig1.pdf+ and (2) \Verb+fig1.pdf+ was created in
\Verb+fig1.R+. In a future where R still exists, changing the figure
will require quick edits of commands already written.  In a future
where R does not exist, I will at least be able to read the plain text
R commands and use them to write code in my new favorite statistical
computing language: R scripts are written in
\href{http://en.wikipedia.org/wiki/Plain_text}{plain text}, and plain
text is a format that will be around as long as computer programmers
write computer programs. \footnote{What is more, since R is open
  source, I will be able to download an old version of R, download an
  old fashioned open-source operating system (like Ubuntu 10), and
  run the old-fashioned statistical computing environment in the
  old-fashioned operating system in a virtual machine on my
  new-fashioned actual machine.}

Moreover, realize that file names send messages to your future
self. Name your files with evocative and descriptive names. Your
collaborators are less likely to call you at midnight asking for help
if your files are named \Verb+inequality-and-protest-figures.R+ than
if your files are called \Verb+temp9+ or
\Verb+supercalifragilisticexpialidocious+. The use of the extension
\Verb+.R+ tells us and the operating system that the file contains R
commands. This part of the filename enables us to quickly search our
antique hard drives for files containing R scripts.

\paragraph{Step 1} Know the provenance of your results so that your
future self or current collaborators can quickly and easily reproduce
and thus change your work.% --- to know what
% commands created which figures and tables.

\section{No data analyst is an island for long.}
Data analysis involves a long series of decisions. Each decision
requires justification.  Some decisions will be too small and
technical for inclusion in the published article itself. These need to
be documented in the code itself \citep{nagler1995coding}. Paragraphs
and citations in the publication will justify the most important
decisions. So, one must code to communicate with yourself and
others. There are two main ways to avoid forgetting the reasons you
did something with data: comment your code and tightly link your code
with your writing.\footnote{One can also try the R command
  \Verb+put-it-in-my-mind(reason,importance='high')+ to firmly place a
  reason for a decision into the mind of the analyst. I myself have
  not had much luck with this function.}

\subsection{Code to communicate: Comment your code.}
Comments --- unexecuted text inside of a script --- are a message to
collaborators (including your future self) and other consumers of your
work. In the above code chunk, I used comments to explain the lines to
readers unfamiliar with R and to remember that I had tried a different
specification but decided not to use it because adding the squared
term did not really change the substantive story arising from the
model.\footnote{R considers text marked with \Verb+#+ as a comment.}
Messages left for your future self (or near-future others) help
retrace and justify your decisions as the work moves from seminar
paper to conference paper to poster back to paper to dissertation and
onwards.

@@

% <<reg1,echo=TRUE,results=verb>>=
% ##  Repeat the regression run for Table~\ref{tab:protest}
% ##  Do political rights predict protest reporting?
% lm1 <- lm(protac00 ~ gini04 + meanpr, data=good.df)
% ##  Q: Was it worthwhile controlling for political rights?
% lm2 <- lm(protac00~gini04,data=good.df)
% coef(lm2)[2]-coef(lm1)[2]
% ##  A: Yes (around 3/4 an act of protest difference)
% @

Notice one another benefit of coding for an audience: we learn by
teaching. By assuming that others will look at your code, you will be
more likely to write clearer code, or perhaps even to learn more about
what you are doing as you do it.

Comment liberally. Comments are discarded when R runs analysis or
\LaTeX~turns plain text into page images, so only those who dig into
your work will see them.

\subsection{Code to communicate: Literate programming.}

Imagine you discover something new (or confirm something old). You
produce a nice little report on your work for use in discussions of
your working group or as a memo for a web or reviewer appendix. The
report itself is a pdf file or some other format which emphasizes
reading. Eventually pieces of that report (tables, graphs, paragraphs)
ought to contribute to the publishable paper. Re-create those analyses
by pointing, clicking, copying, or pasting would invite typing error
and waste time. Re-creating your arguments justifying why you dropped
China but not India from your analyses or why you used a quadratic
term but not a cubic term would also waste time. More importantly, we
and others want to know why we did what we did. Such explanations may
not be very clear if we have some pages of printed code in one hand
and a manuscript in the other.

How might one avoid these problems?
\href{http://en.wikipedia.org/wiki/Literate_programming}{Literate
  programming} is the practice of weaving code into a document ---
paragraphs, equations, and diagrams can explain the code, and the code
can numbers, figures, and tables (and diagrams and even equations and
paragraphs). Literate programming is not merely fancy commenting but
is about enabling the practice of programming itself to enable easy
reproduction and communication.

For example, in \S\ref{sec:data-analys-comp}, I suggested that we know
where ``that plot on page 10'' comes from by making sure we had a
\Verb+fig1.pdf+ file produced from a clearly commented plain text file
called something like \Verb+fig1.R+.  An even easier solution would be to
directly include a chunk of code to produce the figure inside of the
paper itself. This paper, for example, was written in plain text using
\LaTeX~ markup with R code chunks to make things like
Figure~\ref{fig:giniprot}: this combination of \LaTeX and R is called
Sweave \citep{Leis:2005}.\footnote{Support for Sweave is included with
  R.}

\begin{Verbatim}[fontsize=\footnotesize]
This paper, for example, was written in plain text using \LaTeX
markup with R code chunks to make things like
Figure~\ref{fig:giniprot}: this combination of \LaTeX and R is
called Sweave \citep{Leis:2005}.\footnote{Support for Sweave is
  included with R.}
\begin{figure}[H]
\centering
 <<fig1plot,echo=FALSE,fig=TRUE>>=
 ##  Make a scatterplot of Protest by Inequality
 with(good.df,plot(gini04,protac00,xlab='Gini Coefficient 2004 (UNDP)',
                   ylab='Mean Protest Activities\n(World Values Survey 1980-2000)'))
 ##  Label a few interesting points
 with(good.df[c("EGY","JOR","USA","SWE","CHL"),],
      text(gini04,protac00,labels=Nation))
 @
\caption{Protest activity by income inequality \citep[from][]{norris2009data}.}
\label{fig:giniprot}
\end{figure}
\end{Verbatim}

<<fig1code,echo=FALSE,fig=FALSE>>=
##  Make a scatterplot of Protest by Inequality
with(good.df,plot(gini04,protac00,
                  xlab='Gini Coefficient 2004 (UNDP)',
                  ylab='Mean Protest Activities\n(World Values Survey 1980-2000)',
                  cex=.8))

##  Label a few interesting points
with(good.df[c("EGY","JOR","USA","SWE","CHL"),],
     text(gini04,protac00,labels=Nation,srt=0,cex=.6,pos=3,offset=.1))
@

\begin{figure}[H]
  \begin{center}
<<fig1plot,echo=FALSE,fig=TRUE,width=4,height=4>>=
par(bty="n",xpd=TRUE,pty="s",tcl=-.25)
<<fig1code>>
@
    \caption{Protest activity by income inequality \citep[from][]{norris2009data}.}
    \label{fig:giniprot}
  \end{center}
\end{figure}

By using \Verb+\label{fig:giniprot}+, I do not need to keep track of
the figure number. Nor do I need a separate \Verb+fig1.R+ file or
\Verb+fig1.pdf+ file.  Tables and other numerical results are also
possible to generate within in the source code of a
scholarly paper. Those who view the code for this essay will see how
Table~\ref{tab:protest} was also generated directly from a regression
object.\footnote{\citet{beck2010reg} inspired this
  particular presentation of a linear model.}

<<maketable, echo=FALSE, results=hide>>=
  lm1<-lm(protac00~gini04+meanpr,data=good.df)  #  Run the regression

@@

library(xtable)
thextab<-xtable(themat,align=c("l",rep(".",4)),digits=1,
                caption="People living in countries with unequal
  income distributions report less protest activity to World Values
  Survey interviewers than people living in countries with relatively
  more equal income distributions, adjusting for average political
  rights as measured by Freedom House 1980--2000. Data from \\citep{norris2009data}.",label="tab:protest")
@

<<printtable,echo=FALSE,results=tex>>=
  print(thextab,include.colnames=FALSE,
        hline.after=NULL,
        table.placement="H",
        add.to.row=list(
          pos=list(0,3),
          command=c(
            "\\toprule \n &
             \\multicolumn{1}{c}{Coef} &
             \\multicolumn{1}{c}{Std. Err.} &
             \\multicolumn{2}{c}{95\\% CI} \\\\ \\cmidrule(r){2-5} ",
            paste("& \\multicolumn{4}{c}{n: ",n,",",
                  "   resid.sd: ",signif(est.sigma),
                  ", R$^2$: ",r2,"} \\\\ \\bottomrule",sep="")
            )
          ))
@

Literate data analysis is not the same as Sweave, even if Sweave is a
nice implementation.\footnote{The R project has a task view devoted to
  \href{http://cran.r-project.org/web/views/ReproducibleResearch.html}{reproducible
    research} listing many of the different approaches to literate
  programming for R.} If your workflow does not involve \LaTeX~ and R,
you can still implement some of the principles
here. \href{http://www.lyx.org/}{LyX} offers a WYSIWYG environment for
\LaTeX~ which supports Sweave. And the \Verb+odfWeave+ package in R
allows the use of OpenOffice documents in exactly the same
way.\footnote{A quick Google search of ``Sweave for Stata'' turned up
  lots of resources for literate programming with Stata.} Something
other than than Sweave will make our lives easier in the future. Then
we'll change.

\paragraph{Step \thesection}

\begin{quote}
  Let us change our traditional attitude to the construction of
  programs: Instead of imagining that our main task is to instruct a
  computer what to do, let us concentrate rather on explaining to
  human beings what we want a computer to do.
  \citep[][p. 97]{knuth1984literate}
\end{quote}

If we focus on explaining our data analysis to human beings, we will
do a better job with the data analysis itself: we will learn as we
focus on teaching, and we will avoid errors and wasting time as we
ensure that others (including our future selves) can retrace our
steps.

Keep in mind the distinction between the
``source code'' of a document (i.e. what computation was required to
produce it) and the visible, type-set page image. Page images are
great for reading, but not great for reproducing or collaborating. The
source code of any document exchanged by the group must be available
and executable.

@@

\paragraph{Step \thesection} Comments and other documentation enable you to
learn through teaching, code efficiently, and avoid duplicating your
effort. Literate programming also enables you to avoid the kinds of
errors which occur when you re-type numbers in to tables or
graphic-producing systems or even into a document itself. Finally, and
most importantly, a document that can be ``run'' to reproduce all of
the results is a document that can more effectively spur discussion
and learning and cumulation of research as people no longer need spend
weeks attempting to reproduce research on which they desire to build.

@@

\paragraph{Step \thesection} We should know data where the data came from and
what operations were performed on which set of data.

@@

\paragraph{Step \thesection} Writing is rewriting. Thus, all writing involves
versions. When we collaborate with ourselves and others we want to
avoid clobbering and we want to enable graceful reconciliation of
rewriting. One can do these things with formal systems of software
(like subversion, git, etc...) or with formal systems of file naming,
file comparing, and communication or, even better, with both.

@@

\paragraph{Step \thesection} You cannot forsee all of the ways that your code
could be used, but you can at least make sure it does what it is
supposed to do in your particular case.

@@

We all always collaborate. Many of us collaborate with groups of
people at one moment in time as we race against a deadline. All of us
collaborate with ourselves over time.\footnote{What is a reasonable
  time-span for which to plan for self-collaboration on a single idea?
  Ask your advisers how long it took them from idea to dissertation to
  publication.}  The time-frames over which collaboration are required
--- whether among a group of people working together or within a
single scholar's productive life or probably both --- are much longer
than any given version of any given software will easily exist. Plain
text is the exception. Thus, even as we extol version control systems,
one must have a way to ensure future access to them in a form that
will still be around when sentient cockroaches finally join political
science departments (by then dominated by cetaceans after humans are
mostly uploads).\footnote{The arrival of the six-legged social
  scientists revives Emacs and finally makes Ctrl-a Ctrl-x Esc-x
  Ctrl-c a \href{http://kieran.healy.usesthis.com/}{reasonable key
    combination}.}
